---
title: "breast_eQTMs"
author: "Niyati Jain"
date: "2023-07-17"
output: html_document
---

```{r}
library(tidyverse)
library(data.table)
library(readxl)
library(ggplot2)
library(sva)
library(purrr)
```

Expression and DNAm and Covariates
Change the input data for lung, prostate, and colon 
Expression: colon_filtered_genes_exp.RData, lung_filtered_genes_exp.RData, prostate_filtered_genes_exp.RData
DNAm: noob_final_BMIQ_colon_2-6-2021.RData,noob_final_BMIQ_lung_2-6-2021.RData, noob_final_BMIQ_prostate_2-6-2021.RData
pairs: colon_DNAm_Expr_FDR05.txt, lung_DNAm_Expr_FDR05.txt, prostate_DNAm_Expr_FDR05.txt
Covariates: cov_colon_for_age_analysis.txt, cov_lung_for_age_analysis.txt, cov_prostate_for_age_analysis.txt
Episcore: colon_episcore_estimates.txt, lung_episcore_estimates.txt, prostate_episcore_estimates.txt

Example with BREAST
```{r}
#Expression data
expr <- load("/Volumes/groups/Projects/GTEx/Age_TL_EWAS/Age_TL_expression_assoc/expression_data/breast_filtered_genes_exp.RData")
#DNAm data
dnam <- load("/Volumes/groups/Projects/GTEx/Age_TL_EWAS/Lin_analysis/noob_final_BMIQ_breast_2-6-2021.RData")
#Covariates for the DNAm analysis
cov <- read.table("/Volumes/groups/Projects/GTEx/Age_TL_EWAS/Lin_analysis/cov/age_cov/cov_breast_for_age_analysis.txt", sep = "\t", header=T)
#Table containing age-related CpGs (FDR 0.05), the associated gene expression results ("generated in the eQTM_input_file.Rmd script)
pairs <- read.table("/Volumes/groups/Projects/GTEx/niyati/Aging_EWAS_new_SVA/DNAm-Expression/Data/breast_DNAm_Expr_FDR05.txt", sep = "\t", header=T)
#Episcore cell type estimates for breast tissue
ctc <- read.table("/Volumes/groups/Projects/GTEx/niyati/Aging_EWAS_new_SVA/Episcore/Data/breast_episcore_estimates.txt", sep = "\t", header = T)
```


```{r}
#format the colnames of gene expression data 
new_colnames <- gsub("\\.", "-", colnames(breast_filtered_gene_exp))  # Replace "." with "-"
new_colnames <- gsub(".{6}$", "", new_colnames)         # Remove last 6 characters
colnames(breast_filtered_gene_exp) <- new_colnames
rownames(breast_filtered_gene_exp) <- breast_filtered_gene_exp[, 1]
breast_filtered_gene_exp <- breast_filtered_gene_exp[, -1]

row_values_2 <- as.data.frame(cov$CollaboratorSampleID)
colvalues <- as.data.frame(colnames(breast_filtered_gene_exp))

colnames(row_values_2) <- c("Name")
colnames(colvalues) <- c("Name")

#extract the sample IDs for samples where we have DNAm data 
overlap <- row_values_2 %>% inner_join(colvalues) 
sample <- overlap$Name
selected_samples<- breast_filtered_gene_exp[, sample]

#select the genes, where we have CpG-gene pairs with DNAm and gene expression data
geneIds <- pairs$Ensembl
selected_samples_genes <- selected_samples[geneIds, ]
```

```{r}
#select the CpGs, where we have CpG-gene pairs with DNAm and gene expression data
row_values <- pairs$CpG
selected_CpGs <- noob_final_BMIQ_breast[row_values, ]
selected_CpGs_samples <- selected_CpGs[,sample]
```

```{r}
#transform the data
selected_CpGs_samples_T = t(selected_CpGs_samples)
selected_genes_samples_T = t(selected_samples_genes)
```

```{r}
#format the covariates file
colnames(ctc)[1] <- "CollaboratorSampleID"
cov <- inner_join(cov, ctc, by='CollaboratorSampleID')
covariates <- cov[match(sample, cov$CollaboratorSampleID), ]
```

```{r}
#Regression analysis function
#For each pairs of CpG and gene, extract the DNAm and gene expression data, respectively
#combine the data with covariates and run the lm model 
regression_table <- map2_df(pairs$CpG, pairs$Ensembl, function(x_val, y_val) {
  selected_cols1 <- selected_CpGs_samples_T[, x_val]
  selected_cols2 <- selected_genes_samples_T[, y_val]
  combined_data <- cbind(selected_cols1, selected_cols2, covariates)
  model <- lm(selected_cols2 ~ selected_cols1 + BMI + TRISCHD + as.factor(sample_group) + as.factor(raceRE) + Basal + EC + Fat + Fib + Luminal + Lym + MP, data = combined_data)
  p_value <- summary(model)$coefficients[2, "Pr(>|t|)"]
  estimate <- summary(model)$coefficients[2, "Estimate"]
  
  data.frame(CpG = x_val, Ensembl_ID = y_val, p_value = p_value, estimate = estimate, stringsAsFactors = FALSE)
})

#format the result table
regression_table_final <- regression_table %>% mutate(adjusted = p.adjust(p_value, method = "fdr", n = length(p_value))) %>% inner_join(pairs, by = 'CpG', relationship = "many-to-many") %>% select(CpG, Ensembl, p_value, estimate, adjusted, gene) 
regression_table_final <- regression_table_final %>% mutate(id = paste0(CpG,"_",Ensembl)) %>% filter(!duplicated(id))
```

